Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-12 bbbq_article.tex
Describe SNP results better

The first query is a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism Homo sapiens.
This results in \richel{1126} gene IDs.
The next query is a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
SNPs associated with the gene name. The number of SNPs
was limited to the first \richel{100} SNPs per gene,
resulting in \richel{112600} SNPs 
\richel{Note that \cite{musumeci2010single} demonstrates for a way to get
try SNPs, instead of SNDs, maybe I should use that}
.
Per selected SNP, all variation (i.e. descriptions of the changes on DNA
and/or RNA and/or protein level) are obtained from the 'dbSNP' database
as well. Of these variations, only the variations that result in a single
amino acid substitution are kept, resulting in \richel{???} protein changes.
Per protein substitution (i.e. a protein name and an index of the changed residue), 
the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology is determined 
using PureseqTM.
From the topology and the known location of the amino acid substitution, 
it is tallied if the change occurred within a TMH or not.
Also, the percentage of the topology that is TMH is measured.
Not all membrane proteins have a transmembrane helix, resulting
in \richel{???} proteins with at least one TMH.

@@

SNPs occur more often in soluble than in transmembrane protein regions,
as shown in figure \ref{fig:f_snps_found_and_expected}.

>>>>

\paragraph{Pipeline}

The first query is a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism Homo sapiens.
This results in 1129 gene IDs (note that this number varies through time).
The next query is a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
variations associated with the gene name. 
The number of variations
was limited to the first 250 variations per gene,
resulting in 282250 variations. A variation need not always be a SNP,
as dbSNP also catalogs other DNA alterations, such as, among others, insertions,
deletions and tandem repeats.
We select only the variations that result in a SNP for
a single amino acid substitution.
Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology is determined 
using PureseqTM.
From the topology and the known location of the SNP, 
we score the location (i.e. transmembrane or soluble protein region) 
where the change occurred.

@@

We obtained 1129 gene names associated with the key phrase 'membrane protein'.
These genes are linked to 5203 proteins, of which 2754
are predicted to be transmembrane proteins.
We obtained 4811 variations that resulted in an
amino acids substitution, of which 2568 were located 
in predicted transmembrane proteins.
Per protein, we predicted the percentage of TMH
and calculated the percentage of amino acids substitutions
occurring in transmembrane regions.
These points, one per protein, were plotted as depicted in figure
\ref{fig:f_snps_found_and_expected}.
We did a linear fit on the data, with and without
membrane-associated proteins, and added a 95\%
confidence interval.
Also, we added the line \inlineequation{y = x},
which indicates the linear fit that would be expected if
SNPs occur just as likely within transmembrane and soluble
protein domains. Already these trend lines
hint that variations occur more often in soluble 
than in transmembrane protein regions.

To test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution. We first calculated the expected number of SNPs
found in TMHs, which is 61457 SNPs multiplied by 6.8\% 
of all proteins being TMH, resulting in the null hypothesis
of 4203 SNPs being in the TMH. 
Note that including or excluding
the membrane associated proteins 
has no effect on this statistical test, as the chance 
of a successful trial is zero (there is zero chance
to find a SNP in TMHs if there are none).
The actually number of SNPs found in TMHs
was 3889, which is lower than the null hypothesis. 
The chance to find, within TMHs, this amount or less SNPs 
equals $1.6201 \cdot 10^{-9}$.
Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we concluded that there are significantly less SNPs
in TMHs as expected by chance.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

We decided to split this analysis based on the number of TMHs
a protein has. We hypothesize that single-spanners (i.e. proteins
with one TMH predicted) are less conserved, when compared to multi-spanners,
as single spanners just need to span a membrane, where some multi-spanners
have enzymatically active sites embedded within their TMHs and thus
have an additional structural constraint.
As can be seen in figure \ref{fig:f_snps_found_and_expected_per_n_tmhs}, 
there is no clear pattern to be observed.
